---
title: "tutorial_data_intro"
author: "Mir Henglin"
date: "6/25/2019"
output: html_document
---

To construct out plots in R, we will be using the `ggplot2` package. To perform data manipulation, we will be using the `dplyr` package.

```{r library, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
```

```{r import, include = FALSE, message=FALSE, warning=FALSE}
plot_data <-
  readRDS('data/tutorial_plot_data.rds') 
```

## Data Format

Rainplots are useful for summarizing the results of multiple models at the same time. In order to plot those results in `ggplot2`, those results must formatted properly. Specifically, the data must be organized in a `data.frame` with columns indicating:

* The model that the results came from
* The term that was evaluated
* The P-value
* The regression estimate

An example of data organized in this way can be seen below. Other columns can be included in the data, but for the purpose of this tutorial we are focusing on the four necessary columns.

```{r}
plot_data
```

In this dataset, `estimate` and `p.value` indicate the regression estimate and the p.value of that estimate. `term` indicates the ID of the metabolite included in the model. `response` indicates which model that a term corresponds to. For example if `response` = Body Mass Index, this indicates that the regression estimate and the p.value correspond to the model where Body Mass Index was the response variable.
